---
title: "mtDNA-CN analysis"
author: "Siyu Pan"
date: "2024/4/5"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn =-1)
```
# 1.Introduction
Mitochondria are essential organelles in human cells, playing pivotal roles in cellular processes such as respiration, apoptosis regulation, and lipid metabolism. Cells contain a variable number of mitochondria. The total number of copies of mtDNA per cell (mtDNA-CN) varies widely depending on multiple factors, such as age, gender and tissue. This study utilizes simulated data to explore and delineate the differences in mtDNA-CN across these diverse factors, offering insights into the complex dynamics of mitochondrial genetics.

# 2.Data Simulation
```{r data-simulation}
# Non-normally distributed data of mtDNA-CN were constructed by combinatorial effects. mtDNA-CN was affected by age, gender, initial DNA concentration, and sequencing plates (random effect). Among them, females are expected to be more strongly affected by age in Muscle.
# Setting seeds to ensure reproducible results.
set.seed(123)
# Sample size
n <- 2000
# Age
ages <- sample(18:80, n, replace = TRUE)
# Gender
gender <- sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.5, 0.5))
# Group
groups <- sample(c("Muscle", "Liver"), n, replace = TRUE, prob = c(0.5, 0.5))
# Sequencing plate
plates <- factor(sample(1:6, n, replace = TRUE))
# DNA concentration, Including some missing values with 3%
dna_conc <- runif(n, 2, 6)
# The basic effect by creating a linear model
# dna_conc, initial DNA concentration
base_effect <- sample(300:400,1,replace = T) - 0.01 * ages + 0.2 * dna_conc
# Gender effect of Female
gender_effect <- ifelse(gender == 'Female', 12, 0)
# A big age effect, only among women in Muscle
age_effect <- ifelse(groups == 'Muscle' & gender == 'Female', -0.3 * ages, 0)
# Adding additional effects of elder.
additional_age_effect <- ifelse(ages > 55, -0.6 * (ages - 55), 0)
# Adding random effects and noise.
plate_effect <- rnorm(length(levels(plates)), 0, 5)
noise <- rnorm(n, 0, 5)
# Creating complex distributions for mtDNA_CN. 
mtDNA_CN <- base_effect + gender_effect + age_effect + plate_effect[plates] + additional_age_effect + noise
data <- data.frame(Group = groups,mtDNA_CN = mtDNA_CN, Age = ages, Gender = gender, Plate = plates, DNAConc = dna_conc)
```

# 3.Data Processing
```{r Data processing}

# Removing outliers other than the 3sd differences
CN_mean <- mean(data$mtDNA_CN)
CN_sd <- sd(data$mtDNA_CN)
data <- data[data$mtDNA_CN >= CN_mean-3*CN_sd & data$mtDNA_CN <= CN_mean+3*CN_sd,]
# dummy variates, plates
data$Plate <- as.factor(data$Plate)
dummy_vars_plate <- model.matrix(~ Plate - 1, data = data)
data <- cbind(data, dummy_vars_plate)
data$Plate6 <- NULL
head(data)
```

# 4.Data Analysis
```{r Statistical test - 1.Normal distribution test}
# 1. Normal distribution test
# Failure to meet the normal distribution test (P = 5.47e-06).
shapiro.test(data$mtDNA_CN)
# Data transformation to normal distribution, such as log transformation or Inverse Normal Transformation
# Inverse Normal Transformation
get_inv_norm <- function(x){
  x.qnorm <- rep(NA, length(x))
  x.qnorm[!is.na(x)] <- qnorm((rank(x[!is.na(x)]) - 0.5) / length(x[!is.na(x)]))
  x.qnorm
}
data$mtDNA_CN_INT<- get_inv_norm(data$mtDNA_CN)
# Check the distribution
shapiro.test(data$mtDNA_CN_INT)
```

```{r Statistical test - 2.correlation anaysis and adjust technique effects}
# Correlation implied that corrections required for technical effects including plates and DNA concentration 
suppressPackageStartupMessages(library(corrplot))
# Male coded as 1, Female coded as 2
data$Gender <- ifelse(data$Gender == "Male",1,2)
# Correlation matrix
cor_matrix <- cor(subset(data, select = -c(Group,Plate)), method = "pearson")
# Heat plot
corrplot(cor_matrix, method = 'square',sig.level = 0.05,insig = "label_sig", pch.cex = 0.9, pch.col = 'grey20',col=rev(COL2('RdBu', 200)))

# Adjusted technique effects by creating a mixed linear model where Plates are used as a random effect
suppressPackageStartupMessages(library(lme4))
model_lmm <- lmer(mtDNA_CN_INT ~  DNAConc + (1 | Plate), data = data)
summary(model_lmm)
# Obstaining the residual
data$mtDNA_CN_INT_residual <- residuals(model_lmm)
```

```{r Statistical test - 3.compare mtDNA-CN differences in different tissues}
# t-test or wilcox test was used to compare mtDNA-CN differences between Muscle and Liver.
# Muscle has significantly more mtDNA-CN than Liver with t test and wilcox test.
t.test(mtDNA_CN_INT_residual ~ Group, data = data)
wilcox.test(mtDNA_CN_INT_residual ~ Group, data = data)
# Density Plot of mtDNA-CN by Group (tissue)
suppressPackageStartupMessages(library(ggplot2))
ggplot(data, aes(x = mtDNA_CN_INT_residual, fill = Group)) + 
  geom_density(alpha = 0.5) +
  theme_bw() +
  labs(title = "Density Plot of mtDNA-CN by Group",
       x = "mtDNA-CN",
       y = "Density") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_fill_brewer(palette = "Set1") 
# Density Plot of mtDNA-CN by Gender
ggplot(data, aes(x = mtDNA_CN_INT_residual, fill = factor(Gender))) + 
  geom_density(alpha = 0.5) +
  theme_bw() +
  labs(title = "Density Plot of mtDNA-CN by Gender",
       x = "mtDNA-CN",
       y = "Density") +
  theme(plot.title = element_text(hjust = 0.5)) +
   scale_fill_brewer(palette = "Set1", name = "Gender", labels = c("Male", "Female"))
```

```{r Statistical test - 4.compare mtDNA-CN differences in different gender}
# Differences in mtDNA-CN in different gender by group
# Compared to muscle, mtDNA-CN was significantly higher in females than males in Liver using t test and wilcox test
t.test(mtDNA_CN_INT_residual ~ Gender, data = data[data$Group == "Liver",])
wilcox.test(mtDNA_CN_INT_residual ~ Gender, data = data[data$Group == "Liver",])
t.test(mtDNA_CN_INT_residual ~ Gender, data = data[data$Group == "Muscle",])
wilcox.test(mtDNA_CN_INT_residual ~ Gender, data = data[data$Group == "Muscle",])
# Faceted box plots by group and gender 
ggplot(data = data, aes(x = Group)) +
  geom_boxplot(alpha=0.8, aes(y=mtDNA_CN_INT_residual, fill=factor(Gender))) +
  theme_bw() + 
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) + 
  xlab("Tissue") + ylab("mtDNA-CN") + labs(fill = "Gender") + scale_fill_discrete(labels=c("Male","Female"))
```

```{r Statistical test - 5.Relationship between age and mtDNA-CN}

# Relationship between age and mtDNA-CN,some possible nonlinear relationships were observed.
model <- lm(mtDNA_CN_INT_residual ~ Age,data=data)
# Age had a strong effect on mtDNA-CN, explaining the greater mtDNA-CN variance
summary(model)
# Looking for knots
suppressPackageStartupMessages(library(earth))
suppressPackageStartupMessages(library(splines))
mars_model <- earth(mtDNA_CN_INT_residual ~ Age, data=data)
# The possible knot is 54
summary(mars_model)
spline_model <- lm(mtDNA_CN_INT_residual ~ bs(Age, knots=54), data=data)
# Obtaining predicted mtDNA
predict_cn<- predict(spline_model,list(Age=sort(data$Age)),se=T)
# Scatter plot and fitted lines
plot(data$Age, data$mtDNA_CN_INT_residual, main="Linear Spline Model", xlab="Age", ylab="mtDNA-CN")
# Liner mode
abline(model, col="green", lwd=2)
# Linear Spline Model
lines(sort(data$Age), predict_cn$fit, col="red",lwd=2)
```
```{r Statistical test - 6. Relationship between age and mtDNA-CN under multiple feactors}
suppressPackageStartupMessages(library(ggpmisc))
data$Gender <- factor(data$Gender, levels = c(1, 2), labels = c("Male", "Female"))
ggplot(data, aes(x = Age, y = mtDNA_CN_INT_residual)) +
  geom_point() +  
  theme_bw() + 
  geom_smooth(method = "lm",formula = y~x, se = FALSE, color = "blue") +
  facet_grid(Group ~ Gender) +
  stat_poly_eq(aes(label = paste(after_stat(eq.label),sep = "~~~")), formula = y~x, parse = TRUE,size = 3)  + 
  stat_fit_glance(method = 'lm',method.args = list(formula = y ~ x), geom = 'text',aes(label = paste("P-value = ", signif(after_stat(p.value), digits = 2), sep = "")), label.x.npc = 'right', size = 3) + 
  labs(title = "Age vs mtDNA-CN by Group and Gender",y = "mtDNA-CN")
```

# 4.Conclusions
From the analysis of the simulated data, we observe that:
1. Mean mtDNA-CN was significantly higher in females than in males, highlighting women may have a greater need for energy.
2. Mean mtDNA-CN in liver tissue is considerably higher than in muscle tissue, potentially indicating varying mtDNA copy number requirements across different types of tissues.
3. mtDNA-CN showed a decreasing trend with age throughout the sample. Notably, mtDNA-CN declined more rapidly after age 54, suggesting a concern for the body's energy metabolism.
4. The age effect was greater in Muscle in females, and changes in mtDNA-CN were more significantly with age, indicating that both age and gender may collaboratively affect mtDNA-CN levels, especially in specific tissue types.

